Secure Composition of SPECTRE Mitigations

### 2023-11-02 prisc-spectre/prisc.tex
Merge
Incorporate some Feedback

%% Title information
\title{Secure Composition of SPECTRE Variants}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

Some modern programming languages enjoy strong security guarantees, for example the Rust programming languages has memory safety guarantees given by its compiler performing a semantic analysis pass.
While programmers may expect that these guarantees hold even after translating their program to the target programming language, it has been shown that this expectation is false in the general case~\MK{cit constant time examples}.
Even correct compilers do not provide satisfactory guarantees and thus one has to resort to secure compilers~\cite{abate19,abate20,patrignani21} which preserve property satisfaction even when the compiled program is linked with arbitrary target-level code.

@@

To prove individual compilation passes secure and compose them into one compiler, there is a semantic interpretation of the modifications done by a compilation pass~\cite{kruse23,abate21}. 
To instantiate their framework, one has to define a cross-language trace relation that witnesses the behavioral changes of a program when compiled.
It is worth noting that some compiler compositions are not useful in practice, such as one where the last compilation pass invalidates the security guarantees given by other passes, like a pass that inserts accesses to some arbitrary memory address without any checks, thereby violating memory-safety.
Whether a composition is useful or not can be captured by compatibility of the corresponding cross-language trace relation, thus bypassing the need to reason about the concrete, syntactic changes a compilation pass does.

@@

Within the framework of recent work~\cite{kruse22,kruse23}, this extended abstract aims to formalize the semantic effect of mitigation mechanisms for SPECTRE variants and investigate their compatibility with each other.
%
%
%This work aims to investigate the compatibility, within the framework of recent work\MK{kruse et al}, of source-code instrumentation passes that insert SPECTRE mitigations. 
Moreover, the paper investigates not only security-preserving compilers, but also security-\textit{enforcing} compilers. 
Enforcements make no assumptions to the code of a source-level component\footnote{Although the source-level language may impose restrictions by means of a type-system that nonetheless provide security guarantees, since ill-typed programs are not compiled in practice.}.
%% MK:
%% motivation lacking for why we do it for just SPECTRE
%% literally no explanation for spectre whatsoever

@@

In the following, it is assumed that the source-level traces $\src{\trace}$ are trivially absent of speculative leaks. 
This is an acceptable assumption to make, since higher-level programming languages typically do not have a semantics where speculation is considered. 
Consequently, if no speculation happens, no speculative leak can occur.
For the target-level traces $\trg{\trace}$, it is assumed that all vulnerabilities are in some sense expressible in the trace model of the target language. 
At the very least, this paper assumes the existence of some kind of allocation ($\trg{Alloc(\loc;n)}$), memory load/store ($\trg{Get(\loc;idx;n)}$ and $\trg{Set(\loc;idx;n;v)}$), branch ($\trg{If(b)}$), and indirect branch events ($\trg{Ibranch(v)}$) as well as a marker event for a barrier ($\trg{\times}$), and a rollback event ($\trg{Rlb}$)~\cite{patrignani21b}.
The following sketches the semantic effect of popular SPECTRE mitigation techniques and borrows the succinct notation from \gls{ltlp}~\cite{markey03}\footnote{$\square \equiv$ ,,always'', $\lozenge \equiv$ ,,eventually'', $\triangleright \equiv$ ,,next'', and $\triangleleft \equiv$ ,,previous''}.
Note that this paper only considers security tags $\trg{\sigma}$ in the target language, the source language is implicitly secure with respect to speculation, since it does not model any in its dynamic semantics.

@@

  \item \gls{lfence} (\texttt{v1}, \texttt{v4})
$$
\begin{array}{rcl}
  \src{\trace} \sim^{v1,v4}_{\texttt{\tiny LFENCE}} \trg{\trace} & \equiv & \trg{\trace}\vDash\square(\trg{If(\_)^\sigma}\implies\triangleright\trg{\times^{\sigma'}})\\
                                                                 && \text{ and }\trg{\trace}\vDash\square(\trg{Get(\_)^{\sigma''}}\implies\triangleleft\trg{\times^{\sigma'''}})
\end{array}
$$
  \gls{lfence} simply puts a barrier after any branch or before any load instruction. 
  While the literature provides (partial) solutions that do not insert the barrier \textit{everywhere}, due to the significant performance penalty, this pass is simple, stupid and puts the barrier ,,everywhere'', i.e., in front of loads and after branches.
  Since the $\src{S}$-level trace is completely irrelevant, this is an example for an enforcement.

>>>>

%% Title information
\title{Secure Composition of SPECTRE Mitigations}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

Some modern programming languages enjoy strong security guarantees, for example the Rust programming languages has memory safety guarantees given by its compiler performing a semantic analysis pass.
While programmers may expect that these guarantees hold even after translating their program to the target programming language, it has been shown that this expectation is false in the general case~\cite{barthe18}.
Even correct compilers do not provide satisfactory guarantees and thus one has to resort to secure compilers~\cite{abate19,abate20,patrignani21} which preserve property satisfaction even when the compiled program is linked with arbitrary target-level code.

@@

To prove individual compilation passes secure and compose them into one compiler, previous work introduces a framework for a semantic interpretation of the modifications done by a compilation pass~\cite{kruse23,abate21}. 
To instantiate this framework, one has to define a cross-language trace relation that witnesses the behavioral changes of a program when compiled.
It is worth noting that some compiler compositions may not give the wanted security properties, such as one where the last compilation pass invalidates the security guarantees given by other passes, like a pass that inserts accesses to some arbitrary memory address without any checks, thereby violating memory-safety.
Compilation passes that do not violate the security properties of earlier ones fulfill a notion of compatibility of a cross-language trace relation that describes the effect of the compiler. 
Because compatiblity is defined on cross-language trace relations, there is no need to reason about the concrete, syntactic changes a compilation pass does.

@@

Within the framework of recent work~\cite{kruse22,kruse23}, this extended abstract aims to give cross-language trace relations that describe the semantic effect of mitigation mechanisms for SPECTRE variants and investigate their compatibility with each other.
%
%
%This work aims to investigate the compatibility, within the framework of recent work\MK{kruse et al}, of source-code instrumentation passes that insert SPECTRE mitigations. 
Moreover, the paper investigates not only security-preserving compilers, but also security-\textit{enforcing} compilers. 
Enforcements make no assumptions to the code of a source-level component\footnote{Although the source-level language may impose restrictions by means of a type-system that nonetheless provide security guarantees, since ill-typed programs are not compiled in practice.}.
%% MK:
%% motivation lacking for why we do it for just SPECTRE
%% literally no explanation for spectre whatsoever

@@

The following assumes that the source-level traces $\src{\trace}$ are trivially absent of speculative leaks. 
This is an acceptable assumption to make, since higher-level programming languages typically do not have a semantics where speculation is considered. 
Consequently, if no speculation happens, no speculative leak can occur.
For the target-level traces $\trg{\trace}$, it is assumed that all vulnerabilities are in some sense expressible in the trace model of the target language. 
This section also assumes the existence of some kind of allocation ($\trg{Alloc(\loc;n)}$), memory load/store ($\trg{Get(\loc;idx;n)}$ and $\trg{Set(\loc;idx;n;v)}$), branch ($\trg{If(b)}$), and indirect branch events ($\trg{Ibranch(v)}$) as well as a marker event for a barrier ($\trg{\times}$), and a rollback event ($\trg{Rlb}$)~\cite{patrignani21b}.
The following sketches the cross-language trace relations describing the semantic effect of popular SPECTRE mitigation techniques.
Note that this paper only considers security tags $\trg{\sigma}$ in the target language, the source language is implicitly secure with respect to speculation, since it does not model any in its dynamic semantics.

@@

  \item \gls{lfence} (\texttt{v1}, \texttt{v4})
$$
\begin{array}{rcl}
  \src{\trace} \sim^{v1,v4}_{\texttt{\tiny LFENCE}} \trg{\trace} & \equiv & (\text{if }\forall i,\trg{\trace}[i]=\trg{If(\_)^\sigma}\text{ then }\trg{\trace}[i+1]=\trg{\times^{\sigma'}})\\
                                                                 && \text{and }(\text{if }\forall i>0,\trg{\trace}[i]=\trg{Get(\_)^{\sigma''}}\\
                                                                 && \ \ \ \ \ \ \ \text{ then }\trg{\trace}[i-1]=\trg{\times^{\sigma'''}})
\end{array}
$$
  \gls{lfence} simply puts a barrier after any branch or before any load instruction. 
  While the literature provides (partial) solutions that do not insert the barrier \textit{everywhere}, due to the significant performance penalty, this pass is simple, stupid and puts the barrier ,,everywhere'', i.e., in front of loads and after branches.
  Since the $\src{S}$-level trace is completely irrelevant, this is an example for an enforcement.
